---
title: "Raster-vector overlay: tabulation"
author: Insang Song
date: 2024-08-14
format:
  html:
    highlight: pygments
    toc: true
---


# Motivation
- A state-of-the-art raster-vector overlay methods provided in `exactextractr` and `terra` are a bit slow in categorical rasters.
- This notebook demonstrates an alternative way to perform categorical raster frequency tabulation with `collapse` on top of `exactextractr` and `terra`.


# Data
- National Land Cover Database (NLCD) 2021
- US counties in NC 2020
- US Census Block Group in NC 2020


# Approach
- Use `exactextractr` to extract raster values to bare `data.frame` then tabulate with `collapse::qtab()`


# Code

```{r}
#| title: packages

library(sf)
library(terra)
library(exactextractr)
library(collapse)
library(tigris)
options(sf_use_s2 = FALSE)

```


```{r}
#| title: data
options(use_tigris_cache = TRUE)
nccounties <- tigris::counties("NC", year = 2020)
ncbgroups <- tigris::block_groups("NC", year = 2020)

nlcd <- terra::rast(
  file.path("../../..", "Downloads", "nlcd_2021_land_cover_l48_20230630", "nlcd_2021_land_cover_l48_20230630.img")
)

# reproject
nccounties <- st_transform(nccounties, crs = terra::crs(nlcd))
ncbgroups <- st_transform(ncbgroups, crs = terra::crs(nlcd))

# random points
states20 <- tigris::states(year = 2020)
states_excl <- c("02", "15", "60", "66", "68", "69", "72", "78")
states20s <- states20[!states20$STATEFP %in% states_excl, ]
states20s <- st_transform(states20s, crs = terra::crs(nlcd))

states_rps <- st_sample(states20s, 1000)
states_rpsb <- st_buffer(states_rps, 50000)

```


```{r}
#| title: extract with exactextractr

system.time(
    nc_counties_extract <- exact_extract(nlcd, nccounties, fun = "frac")
)
system.time(
    nc_bgroups_extract <- exact_extract(nlcd, ncbgroups, fun = "frac")
)
system.time(
    rpb_extract <- exact_extract(nlcd, states_rpsb, fun = "frac")
)
```


`terra` case is too slow to see the result in this notebook. Therefore, we will skip the case here.

```{r}
#| title: tabulation
#| eval: false

nccountiest <- terra::vect(nccounties)
ncbgroupst <- terra::vect(ncbgroups)
states_rpsbt <- terra::vect(states_rpsb)


system.time(
    nc_counties_terra <- extract(nlcd, nccountiest, fun = table, exact = TRUE)
)
system.time(
    nc_bgroups_terra <- extract(nlcd, ncbgroupst, fun = table, exact = TRUE)
)
system.time(
    rpb_terra <- extract(nlcd, states_rpsbt, fun = table, exact = TRUE)
)

```

The bare `data.frame` is larger than the summary table. It seems like `exactextractr` processes each feature at a time. The last case using buffer polygons crashes even with 32GB memory (10+ GB free).

```{r}
#| title: bare data.frame
#| 

system.time(
    nc_counties_extracto <- exact_extract(nlcd, nccounties)
)
system.time(
    nc_bgroups_extracto <- exact_extract(nlcd, ncbgroups)
)
system.time(
    rpb_extracto <- exact_extract(nlcd, states_rpsb)
)

```


## Row-wise tabulation

```{r}
#| title: row-wise tabulation
library(future)
library(future.apply)
future::plan(future::multicore, workers = 8L)

qex <- function(ras, sf, q = 100, fun = "frac") {
#   qex_res <- vector("list", length = nrow(sf))

  qex_res <- future.apply::future_Map(
    function(i) {
        end <- min(i + q - 1, nrow(sf))
        ex <- exact_extract(ras, sf[i:end, , drop = FALSE])[[1]]
        qex_res[i:end] <- lapply(ex, function(x) collapse::fsum(x$coverage_fraction, x$value))
    }, seq(1, nrow(sf), q))

  return(qex_res)
}

system.time(
    nc_counties_qex <- qex(nlcd, nccounties)
)
system.time(
    nc_bgroups_qex <- qex(nlcd, ncbgroups)
)
system.time(
    rpb_qex <- qex(nlcd, sf::st_as_sf(states_rpsb))
)

``` 


Perhaps the slowness comes from the data size that should be handled by `exactextractr`. Extracting `data.frame` to summarize with `collapse` took longer time than doing summary in `exactextractr`. C++ level parallelization might be beneficial to speed up the process, which is beyond the scope of this document.


## Added 2025-02-25


```{r}
#| title: row-wise mask-freq sequence

library(terra)
library(future)
library(future.apply)

raster_path <- "/mnt/d/chopin-bench/Annual_NLCD_LndCov_2021_CU_C1V0.tif"
gdfbase <- "/mnt/d/chopin-bench/tlgdb_2024_a_us_substategeo.gdb"
raster <- terra::rast(raster_path)
gdf <- terra::vect(gdfbase, layer = "County")
gdf <- terra::project(gdf, terra::crs(raster))
gdfsub <- gdf[grep("^(02|15|68|72|66|69|60|78)", gdf$GEOID, invert = TRUE), ]

future::plan(future::multicore, workers = 8L)

qex <- function(ras, sf) {

  qex_res <- future.apply::future_Map(
    function(i) {
        ex <- terra::crop(ras, sf[i,], touches = TRUE)
        ex <- terra::freq(ex, wide = TRUE)
        return(ex)
    }, seq_len(nrow(sf)))

  return(qex_res)
}

system.time(
    qres <- qex(raster, gdfsub)
)

system.time(
    us_counties_extracto <- exactextractr::exact_extract(raster, sf::st_as_sf(gdfsub), fun = "frac")
)

```